Classless Java

### 2016-03-24 OOPSLA16/sections/CaseStudy.tex
spell and grammar check

This section illustrates applications and larger case studies for
Classless Java. The first application shows how a useful pattern,
using multiple inheritance and type-refinement, can be conviniently
encoded in Classless Java. The second application shows how to model
embedded DSLs based on fluent APIs. The two larger case studies, take
existing projects and refactor them into Classless Java. The first
case study shows a significant reduction in code size, while the
second case study maintains the same amount of code, but improves
modularity.

@@

As a first application for Classless Java, we illustrate a useful
programming pattern that improves modularity and extensibility of
programs. This useful pattern is based on an existing solution to 
the \emph{Expression Problem} (EP)~\cite{wadler98expression}, which is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunately, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing. Our last case study, presented
in Section~\ref{subsec:int}, is essentially a (much larger) application of this
pattern to an existing program. Here we illustrate the pattern in the
much smaller Expression Problem.

@@

\subsection{Refactoring an Interpreter}\label{subsec:int}
The last case study applies refactors the code from an interpreter for
a Lisp-like language
\lstinline{Mumbler}\footnote{https://github.com/cesquivias/mumbler/tree/master/simple},
which is created as a tutorial for the Truffle
Framework~\cite{wurthinger2013one}.  Keeping a balance between
simplicity and useful features, \lstinline{Mumbler} contains numbers,
booleans, lists (encoding function calls and special forms such as
if-expression, lambdas, etc). In the original code base, which
consists of 626 SLOC of Java, only one operation \texttt{eval} is
supported. Extending Mumbler to support one more operation, such as a
pretty printer, would normally require changing the existing code base
directly.

Our refactoring applies the pattern presented in
Section~\ref{subsec:ep} to the existing Mumbler code base to improve
its modularity and extensibility. Using the refactored code base it is 
becomes possible to add new operations modularly, and to support
independent extensibility. We add one more
operation \texttt{print} to both the original and the refactored code
base. In the original code base the pretty printer is added
non-modularly by modifying the existing code. In the refactored code
base the pretty printer is added modularly.
Although the code in the refactored version is slightly increased (by 2.2\% SLOC), the
modularity is greatly increased, allowing for improved reusability and maintainability.

>>>>

This section illustrates applications and larger case studies for
Classless Java. The first application shows how a useful pattern,
using multiple inheritance and type-refinement, can be conveniently
encoded in Classless Java. The second application shows how to model
embedded DSLs based on fluent APIs. The two larger case studies, take
existing projects and refactor them into Classless Java. The first
case study shows a significant reduction in code size, while the
second case study maintains the same amount of code, but improves
modularity.

@@

As the first application for Classless Java, we illustrate a useful
programming pattern that improves modularity and extensibility of
programs. This useful pattern is based on an existing solution to
the \emph{Expression Problem} (EP)~\cite{wadler98expression}, which is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunately, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing. Our last case study, presented
in Section~\ref{subsec:int}, is essentially a (much larger) application of this
pattern to an existing program. Here we illustrate the pattern in the
much smaller Expression Problem.

@@

\subsection{Refactoring an Interpreter}\label{subsec:int}
The last case study refactors the code from an interpreter for
a Lisp-like language
\lstinline{Mumbler}\footnote{https://github.com/cesquivias/mumbler/tree/master/simple},
which is created as a tutorial for the Truffle
Framework~\cite{wurthinger2013one}.  Keeping a balance between
simplicity and useful features, \lstinline{Mumbler} contains numbers,
booleans, lists (encoding function calls and special forms such as
if-expression, lambdas, etc). In the original code base, which
consists of 626 SLOC of Java, only one operation \texttt{eval} is
supported. Extending Mumbler to support one more operation, such as a
pretty printer, would normally require changing the existing code base
directly.

Our refactoring applies the pattern presented in
Section~\ref{subsec:ep} to the existing Mumbler code base to improve
its modularity and extensibility. Using the refactored code base it
becomes possible to add new operations modularly, and to support
independent extensibility. We add one more
operation \texttt{print} to both the original and the refactored code
base. In the original code base the pretty printer is added
non-modularly by modifying the existing code. In the refactored code
base the pretty printer is added modularly.
Although the code in the refactored version is slightly increased (by 2.2\% SLOC), the
modularity is greatly increased, allowing for improved reusability and maintainability.

### 2016-03-24 OOPSLA16/sections/Implementation.tex
spell and grammar check

Thus, in addition to the advantages of Lombok, Classless Java offers
some more advantages with respect to arbitrary (Compilation agent driven) AST rewriting.

@@

Our automatic type refinement is a useful and convenient feature, but
not transparent to the heirs of the annotated interface.  They need to
be aware of the annotation semantics and provide the right type while
refining methods. To support heir coherence, we would need
to give up automatic type refinement, which is an essential part of IB programming.
However, the reader should note that Java libraries almost always break heir
coherence during evolution and still claim backward compatibility (false in
theory but statistically true in practice). In practice, adding any method to any
non final class of a Java library is enough to break heir
coherence.  We think return type refinement breaks heir coherence ``less" than normal library evolution, and
if no automatic type-refinements are needed, then \mixin can claim a
form of heir coherence.

>>>>

Thus, in addition to the advantages of Lombok, Classless Java offers
some more advantages with respect to arbitrary (compilation agent driven) AST rewriting.

@@

Our automatic type refinement is a useful and convenient feature, but
not transparent to the heirs of the annotated interface.  They need to
be aware of the annotation semantics and provide the right type while
refining methods. To support heir coherence, we would need
to give up automatic type refinement, which is an essential part of IB programming.
However, the reader should note that Java libraries almost always break heir
coherence during evolution and still claim backward compatibility (false in
theory but statistically true in practice). In practice, adding any method to any
non-final class of a Java library is enough to break heir
coherence.  We think return type refinement breaks heir coherence ``less" than normal library evolution, and
if no automatic type-refinements are needed, then \mixin can claim a
form of heir coherence.

### 2016-03-24 OOPSLA16/sections/Overview.tex
spell and grammar check

\paragraph{\Q@Point2D@: simple immutable data with fields}
Here we will illustrate how points are modelled with interfaces. In IB
we do not talk about state directly. Instead state is accessed and
manipulated using abstract methods.  The usual approach to model
points in Java is to use a class with fields for the coordinates.
In Classless Java interfaces are used instead:

@@

\paragraph{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
2D points are mathematical entities, thus we choose immutable data structure to
model them. However animals are real world entities, and when an animal moves,
it is the \emph{same} animal with a different location. We model this with
mutable state.

@@

\paragraph{Summary of operations in Classless Java}
%Method bodies can only refer to other methods;
%Objects are (conceptually) closures and the state is
%composed by captured local variables.}
%Methods interacting with state are called \emph{abstract state operations}.
In summary, object interfaces provide support for
different types of abstract state operations: four field-based state
operations; and functional updaters. Furthermore object interfaces support
direct object intantiation via \texttt{of} factory methods.
Figure~\ref{fig:abstractstate} summarizes the six operations supported
by \mixin. The field-based abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in more detail in Section~\ref{sec:dsls}.

>>>>

\paragraph{\Q@Point2D@: simple immutable data with fields}
Here we will illustrate how points are modelled with interfaces. In IB
we do not talk about state directly. Instead, state is accessed and
manipulated using abstract methods.  The usual approach to model
points in Java is to use a class with fields for the coordinates.
In Classless Java interfaces are used instead:

@@

\paragraph{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
2D points are mathematical entities, thus we choose an immutable data structure to
model them. However animals are real world entities, and when an animal moves,
it is the \emph{same} animal with a different location. We model this with
mutable state.

@@

\paragraph{Summary of operations in Classless Java}
%Method bodies can only refer to other methods;
%Objects are (conceptually) closures and the state is
%composed by captured local variables.}
%Methods interacting with state are called \emph{abstract state operations}.
In summary, object interfaces provide support for
different types of abstract state operations: four field-based state
operations; and functional updaters. Furthermore object interfaces support
direct object instantiation via \texttt{of} factory methods.
Figure~\ref{fig:abstractstate} summarizes the six operations supported
by \mixin. The field-based abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in more detail in Section~\ref{sec:dsls}.
